An Open-Source OpenSim Oculomotor Model for Kinematic and Dynamic Simulation

### 2018-07-09 doc/manuscript.tex
fix fixation controller

\documentclass[11pt,a4paper,draft=false]{report}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{bm,txfonts}       % bm: bold for Greek letters, txfonts -> varmathbb
\usepackage{breqn}              % automatically brake line in equations
\usepackage{cancel}
\usepackage{arydshln}         % matrix dash lines
\usepackage{booktabs}         % tables
\usepackage{color,soul}         % used for \hl{}
\usepackage[inline]{enumitem}   % use to set labels in the enumerate environment
\usepackage{fullpage}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figures/}{./tikz/}}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}
\usepackage{caption}
\captionsetup{labelfont=bf}
\usepackage[caption=false,labelfont=bf]{subfig}
\usepackage{listings}
% \usepackage{algorithm}
% \usepackage{algorithmic}
\usepackage{url}
\usepackage[pdfpagelabels,pdfusetitle,colorlinks=true,pdfborder={0 0
  0}]{hyperref}
\usepackage{natbib}
\bibliographystyle{IEEEtranN}

@@

\newacronym{dofs}{DoFs}{Degrees of Freedom}
\newacronym{dof}{DoF}{Degree of Freedom}
\newacronym{eoms}{EoMs}{Equations of Motion}
% \newacronym{so}{SO}{Static Optimization}
\newacronym{pd}{PD}{Proportional Derivative}
\newacronym{cns}{CNS}{Central Nervous System}
\newacronym{lr}{LR}{Lateral Rectus}
\newacronym{sr}{SR}{Superior Rectus}
\newacronym{ir}{IR}{Inferior Rectus}
\newacronym{mr}{MR}{Medial Rectus}
\newacronym{so}{SO}{Superior Oblique}
\newacronym{io}{IO}{Inferior Oblique}
\newacronym{rmse}{RMSE}{Root Mean Square Error}
\newacronym{em}{EOMs}{Extraocular Muscles}
\newacronym{fl}{F-L}{Force-Length}
\newacronym{fv}{F-V}{Force-Velocity}
\makeglossaries{}

@@

\DeclareMathAlphabet\mathbfcal{OMS}{cmsy}{b}{n} % mathcal bf
\newcommand{\ten}[1]{\mathbfcal{#1}}
\newcommand{\mat}[1]{\bm{#1}}
\renewcommand*{\vec}[1]{\bm{#1}}
\newcommand{\pd}[2]{\frac{\partial#1}{\partial#2}}
\newcommand{\subspace}[1]{\varmathbb{#1}}
\newcommand{\rs}[1]{\subspace{R} (#1)}
\newcommand{\cs}[1]{\subspace{C} (#1)}
\newcommand{\ns}[1]{\subspace{N} (#1)}
\newcommand{\fs}[1]{\subspace{F}}
\newcommand{\ms}[1]{\subspace{M}}
\newcommand{\R}[1]{\mathfrak{R}^{#1}}
\newcommand{\inr}[1]{\in\R{#1}}
\newcommand{\TP}[1]{\mat{T}_{#1}}     % tall projection
\newcommand{\FP}[1]{\mat{F}_{#1}}     % fat projection
\newcommand{\NT}[1]{\mat{N}_{\TP{#1}}}
\newcommand{\N}[1]{\mat{N}_{#1}}
\newcommand{\dNT}[1]{\mat{\dot{N}}_{\TP{#1}}}
\newcommand{\NF}[1]{\mat{N}_{\FP{#1}}}
\newcommand{\dNF}[1]{\mat{\dot{N}}_{\FP{#1}}}
\newcommand{\mc}[1]{\mathcal{#1}}

% mathematical proofs
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{corollary}{Corollary}[chapter]
\newcommand{\corollaryautorefname}{Corollary}
\newtheorem{proposition}{Proposition}[chapter]
\newcommand{\propositionautorefname}{Proposition}
\newtheorem{remark}{Remark}[chapter]
\newcommand{\remarkautorefname}{Remark}
\newtheorem{definition}{Definition}[chapter]
\newcommand{\definitionautorefname}{Definition}
\newcommand{\subfigureautorefname}{Figure} % subfloat
\renewcommand*{\figureautorefname}{Figure}
\renewcommand*{\sectionautorefname}{Section}
\renewcommand*{\subsectionautorefname}{Subsection}
\newcommand{\algorithmautorefname}{Algorithm}
\makeatletter
% \renewcommand\thealgorithm{\thechapter.\arabic{algorithm}}
\@addtoreset{algorithm}{chapter}
\makeatother
\def\equationautorefname~#1\null{Equation #1\null}

% \renewcommand{\baselinestretch}{1.5} % line stretch

@@

Rapid and accurate eye movements are crucial for coordinated direction of
gaze. Studying human eye movement has significant implications for improving our
understanding of the oculomotor system and treating visuomotor disorders. Over
the years, biomechanic simulation has provided an analysis tool of different
human movements, especially gait. This analysis can be extended to visual tasks
by analyzing the mechanisms of ocular motility and providing a realistic ocular
model that can be used to investigate muscle activation patterns during static
fixations and the control and dynamics of various eye movements. This model can
be used to calculate metabolic costs of eye movements and also simulate
different eye disorders, such as different forms of strabismus. Furthermore, it
can be easily integrated into other human body models to analyze the relation
between vestibular system and eye movement by simulating their biomechanics in
three dimensions and investigating the function and neural control of the
\gls{em}.

Eye movements are produced through the activation of six \gls{em}. Clinical
trials have provided a profound knowledge of how the \gls{em} rotate the globe,
the resistive tension to this rotation and the length-tension relationship of
the muscles. Various computational models of the extraocular muscle and orbital
mechanics have been proposed, which provide insight and scientific bases for
oculomotor biomechanics, control of eye movement and binocular
misalignment. These models focus on the realism of muscle behavior and they were
based on the viscoelastic properties and physiological data \gls{em}.

The first 3D biomechanical model was developed by \hl{Robinson (1975)}, who
simplified the formulation by only considering the elasticity of the \gls{em}
and ignoring the dynamics. The model incorporates anatomically realistic muscle
paths and empirical innervation-length-tension relationships. To study the
neural control of rapid saccadic movement, models using anatomical and
mechanical properties of \gls{em} have been developed. They take account that
the actual force is a complex nonlinear function of the muscle length, velocity,
and innervation~\cite{Thelen2003, Millard2013}. Such models have the advantage
of supporting 3D dynamic simulations and have been used to analyze neural
controllers and the pulley hypothesis.

Here, we introduce a detailed 3D biomechanical model, based on the work of
\hl{Iskander} of the human eye which can be used for dynamic simulation. The
model is based on the biomechanical simulator
\texttt{OpenSim}~\cite{Delp2007}. This is an open-source software that can
provide the flexibility of in depth parameterization of the model. The
architecture and dynamic muscle properties are based on physiological and
kinematic measurements of the human eye. The model incorporates an eye-globe,
orbital suspension tissues and six muscles with their connective tissues to test
the passive pulley hypothesis. Furthermore, with this model we are able to
assess the excitation and activation patterns for a variety of targets by
applying a closed-loop fixation controller that drives the model to perform
saccadic movements in a forward dynamic manner. The controller minimizes the
error between the desired trajectory (reference input) and the predicted
movement (calculated output).

@@

We have chosen the passive pulley model for our extraocular model in order to
keep it simple and provide faster simulation speed. \autoref{tab:muscle-path}
shows the positions of muscle pulleys, as well as the origin and insertions
points of the \gls{em} onto the eye, with respect the center of the globe. These
data were acquired from \hl{Iskander 2018} which were based on physiological
measurements, but in our case, they were slightly modified to be located outside
of the globe radius in order to not penetrate the eye globe. Since no position
was documented for the origin of the \gls{so}, a point close to the origins of
the rectus muscles was chosen to match the fiber length in the primary position
of the \gls{so} muscle.

@@

The Millard muscle model~\cite{Millard2013} has been adopted for the modeling of
the \gls{em}, which allows to manually fit force-generation dynamics. The six
\gls{em} were modeled using the rigid tendon assumption that ignores the
elasticity of the tendon. This means that the series element of the muscle model
is not included (the tendon length $l^T$ is equal to the tendon slack length
$l_s^T$). This assumption is valid when the ratio of the tendon length to the
muscle length is less or equal to one, as the in the case of all
\gls{em}. \gls{em} are considered parallel-fibered muscles, so the pennation
angle is zero ($\alpha = 0$). Maximum isometric force $f_o^M$, optimal fiber
length $l_o^M$ and tendon length $l^T$ are presented
in~\autoref{tab:muscle-path} and are based on \hl{Iskander 2018}.

The active and passive \gls{fl} curves for the \gls{em} differ from the ones
that of the skeletal muscles. As shown in \autoref{fig:millard-curves} can be
adapted so as to fit the experimental data available for the gls{lr} muscle. The
following values where used for the active \gls{fl} curve:

@@

The parameters that describe the above relationships were chosen to fit the
curves found in \hl{Iskander 2018}. In the plots below, we try to match the
\gls{fl} relationships at maximum activation of the lateral rectus muscle. This
represents the first part of testing the fidelity of the model. Due to lack of
data describing the other muscles’ \gls{fl} relationships, we used the
parameters found describing the normalized curves of active and passive \gls{fl}
relationships of the gls{lr} for the other \gls{em} as well.

\gls{em} have a higher fraction of fast twitch fibers and thus different
\gls{fv} behavior, due to different structures compared to skeletal muscles. The
default Millard \gls{fv} curve was used for the six \gls{em}, since the behavior
of the selected muscle model depends mainly on the maximum contraction velocity
$v^{\text{max}}$. The maximum muscle contraction velocity is tuned so as to
match the peak velocity of saccadic eye movement 15.7 rad / s. Following this
definition, the maximum muscle contraction velocity is given in optimal fiber
length per seconds and it is thus different for each \gls{em} since their
optimal fiber length is different. Also, because of different structure of
neural control of eye movements, activation and deactivation delays are lower
than in skeletal muscles, at about 5 ms.

@@

Wrapping objects help to simulate the proper dynamics. Using wrapping objects
implemented in the \texttt{OpenSim} allows us to model realistic force
directions as the muscle runs over the eyeball. From the point of origin to the
pulley point, the force exerted by the contraction of the muscles is aligned
along one straight line. But, from the pulley point to the insertion point, the
force is distributed along the surface of a sphere. Two separate wrapping
spheres for the rectus muscles and the oblique muscles were created, to avoid
abnormal changes on the \gls{fl} curve as the eyeball rotates in the three
directions.

>>>>

\documentclass[11pt,a4paper,draft=false]{report}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{bm,txfonts}       % bm: bold for Greek letters, txfonts -> varmathbb
\usepackage{breqn}              % automatically brake line in equations
\usepackage{cancel}
\usepackage{arydshln}         % matrix dash lines
\usepackage{booktabs}         % tables
\usepackage{color,soul}         % used for \hl{}
\usepackage[inline]{enumitem}   % use to set labels in the enumerate environment
\usepackage{fullpage}
\usepackage[pdftex]{graphicx}
\graphicspath{{./figures/}{./tikz/}}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}
\usepackage{caption}
\captionsetup{labelfont=bf}
\usepackage[caption=false,labelfont=bf]{subfig}
\usepackage{listings}
% \usepackage{algorithm}
% \usepackage{algorithmic}
\usepackage{url}
\usepackage[pdfpagelabels,pdfusetitle,colorlinks=true,pdfborder={0 0
  0},pdfauthor={Dimitar Stanev}]{hyperref}
\usepackage{natbib}
\bibliographystyle{IEEEtranN}

@@

\newacronym{dofs}{DoFs}{Degrees of Freedom}
\newacronym{dof}{DoF}{Degree of Freedom}
\newacronym{eoms}{EoMs}{Equations of Motion}
% \newacronym{so}{SO}{Static Optimization}
\newacronym{fd}{FD}{Forward Dynamics}
\newacronym{pd}{PD}{Proportional Derivative}
\newacronym{cns}{CNS}{Central Nervous System}
\newacronym{lr}{LR}{Lateral Rectus}
\newacronym{sr}{SR}{Superior Rectus}
\newacronym{ir}{IR}{Inferior Rectus}
\newacronym{mr}{MR}{Medial Rectus}
\newacronym{so}{SO}{Superior Oblique}
\newacronym{io}{IO}{Inferior Oblique}
\newacronym{rmse}{RMSE}{Root Mean Square Error}
\newacronym{em}{EOMs}{Extraocular Muscles}
\newacronym{fl}{F-L}{Force-Length}
\newacronym{fv}{F-V}{Force-Velocity}
\makeglossaries{}

@@

\DeclareMathAlphabet\mathbfcal{OMS}{cmsy}{b}{n} % mathcal bf
\newcommand{\ten}[1]{\mathbfcal{#1}}
\newcommand{\mat}[1]{\bm{#1}}
\renewcommand*{\vec}[1]{\bm{#1}}
\newcommand{\R}[1]{\mathfrak{R}^{#1}}
\newcommand{\inr}[1]{\in\R{#1}}

\newcommand{\subfigureautorefname}{Figure} % subfloat
\renewcommand*{\figureautorefname}{Figure}
\renewcommand*{\sectionautorefname}{Section}
\renewcommand*{\subsectionautorefname}{Subsection}
\newcommand{\algorithmautorefname}{Algorithm}
\makeatletter
% \renewcommand\thealgorithm{\thechapter.\arabic{algorithm}}
\@addtoreset{algorithm}{chapter}
\makeatother
\def\equationautorefname~#1\null{Equation #1\null}

@@

Rapid and accurate eye movements are crucial for coordinated direction of
gaze. Studying human eye movement has significant implications for improving our
understanding of the oculomotor system and treating vasomotor visuomotor disorders. Over
the past decades, biomechanics simulation has provided the means to analyze
different human movements. The same principles can be used to analyze visual
tasks by modeling the musculoskeletal properties of the oculomotor
system. Consequently, this model can be used to investigate muscle activation
patterns during static fixation, analyze the dynamics of various eye movements,
calculate metabolic costs and simulate different eye disorders, such as
different forms of strabismus. Last but not least, it can be easily integrated
with available full body models in order to analyze the relation between the
vestibular and oculomotor systems.

Eye movements are a result of the coordination of the six \gls{em}. Clinical
trials have provided a profound knowledge of how the \gls{em} act on the eye
globe, the resistive tension of the surrounding tissues and the length-tension
relationship of the muscles. Various computational models of the extraocular
muscles and orbital mechanics have been proposed, which provide insight and
scientific bases for oculomotor biomechanics, control of eye movement and
binocular misalignment. These models focus on the realism of muscle behavior and
they were based on the viscoelastic properties and physiological data \gls{em}.

The first 3D biomechanical model was developed by~\cite{Robinson1964a,
  Robinson1969}, who simplified the formulation by considering the elasticity of
the \gls{em} ignoring their dynamics. The model incorporates anatomically
realistic muscle paths and empirical innervation-length-tension
relationships. To study the neural control of rapid saccadic movement, models
using anatomical and mechanical properties of \gls{em} have been developed. They
take account that the actual force is a complex nonlinear function of the muscle
length, velocity, and innervation~\cite{Thelen2003, Millard2013}. Such models
have the advantage of supporting 3D dynamic simulations and have been used to
analyze neural controllers and the pulley hypothesis.

We introduce a detailed open-source 3D biomechanical model of the human eye,
which can be used for dynamic simulation\footnote{SimTK project:
  \url{https://simtk.org/projects/eye}}, based on the data reported
in~\cite{Iskander2018}. The model is implemented in
\texttt{OpenSim}~\cite{Delp2007}, which is an open-source framework for modeling
and simulation of musculoskeletal systems. The architecture and dynamic muscle
properties are based on physiological and kinematic measurements of the human
eye. The model incorporates an eye-globe, orbital suspension tissues and six
\gls{em}. Using this model we are able to assess the excitation and activation
patterns for a variety of targets by applying a closed-loop fixation controller
that drives the model to perform saccadic movements in a \gls{fd} manner. The
controller minimizes the error between the desired saccadic trajectory
(reference input) and the predicted movement (calculated output).

@@

The model relies on the passive pulley assumption in order to keep it simple and
provide faster simulation speed. \autoref{tab:muscle-path} shows the positions
of muscle pulleys, as well as the origin and insertion points of the \gls{em},
defined in the local body coordinates of the globe. The data is based on
physiological measurements~\cite{Iskander2018}, with some minor adaptation so as
to prevent penetration into the eye globe. Since no position was documented for
the origin of the \gls{so}, a point close to the origins of the rectus muscles
was chosen to match the fiber length in the primary position of the \gls{so}
muscle.

@@

The Millard muscle model~\cite{Millard2013} has been adopted for the modeling of
the \gls{em}, which allows to manually fit the reported data for the \gls{fl}
curves. The muscles were modeled using the rigid tendon assumption that ignores
the elasticity of the tendon. This means that the series element of the muscle
model is not included (the tendon length $l^T$ is equal to the tendon slack
length $l_s^T$). This assumption is valid when the ratio of the tendon length to
the muscle length is less or equal to one, as the in the case of all
\gls{em}. \gls{em} are considered parallel-fibered muscles, so the pennation
angle is zero ($\alpha = 0$). Maximum isometric force $f_o^M$, optimal fiber
length $l_o^M$ and tendon length $l^T$ are presented
in~\autoref{tab:muscle-path}.

The active and passive \gls{fl} curves for the \gls{em} differ from that of a
skeletal muscle. As shown in \autoref{fig:millard-curves}, we can fine-tune
these curves so as to fit the experimental data available for the gls{lr}
muscle. The following values where used for the active \gls{fl} curve:

@@

The parameters that describe the above relationships were chosen to fit the
curves reported in~\cite{Iskander2018}, by matching the \gls{fl} relationship at
maximum activation of the \gls{lr} muscle. This represents the first part of
testing the fidelity of the model. Due to lack of data describing the other
muscles’ \gls{fl} relationships, we used the parameters found describing the
normalized curves of active and passive \gls{fl} relationships of the gls{lr}
for the other \gls{em} as well.

\gls{em} have a higher fraction of fast twitch fibers and thus different
\gls{fv} behavior, due to different structures compared to skeletal
muscles. Despite that, the default Millard \gls{fv} curve was used for the six
\gls{em}, since the behavior of the selected muscle model depends mainly on the
maximum contraction velocity $v^{\text{max}}$. The maximum muscle contraction
velocity is tuned so as to match the peak velocity of saccadic eye movement 15.7
rad / s. Following this definition, the maximum muscle contraction velocity is
given in optimal fiber length per seconds and it is thus different for each
\gls{em}, as their optimal fiber length is different. Furthermore, because of
the different structure of neural control of the eyes, activation and
deactivation delays ($\tau_d = 5$ ms) are lower than in skeletal
muscles. Finally, two separate wrapping spheres for the rectus muscles and the
oblique muscles were created, to avoid abnormal changes on the \gls{fl} curve as
the eyeball rotates in the three directions.

@@
